---
title: "Data Exploration"
output: html_notebook
---

```{r}
library(tidyverse)
library(tidytext)
library(stringr)
library(widyr)

```


```{r}
getwd()
```
```{r}
list.files("texts", full.names = TRUE)
```
```{r}
read_file("texts/A06785.txt")
```


```{r}

files <- list.files("texts",
                    pattern = "\\.txt$",
                    full.names = TRUE)

corpus_raw <- tibble(
  doc_id = tools::file_path_sans_ext(basename(files)),
  text = map_chr(files, read_file)
)

print(corpus_raw)
```

```{r}
# Count words that contain "ſ" before replacement
sum(str_detect(corpus_raw$text, "ſ"))
```


Step 0: Normalization

In preparing the corpus for analysis, I adopted a minimal-intervention 
normalization strategy. The goal was to ensure lexical comparability across 
documents while preserving historically meaningful orthographic variation.


0.1) First, I replaced the Early Modern long S character (ſ) with the modern 
letter “s,” as required. Without this correction, words like “ſtate” and “state”
would be counted as different words, which would distort the analysis.

0.2) Second, I converted all text to lowercase and removed extra spaces. 
Lowercasing ensures that words such as “State” and “state” are treated as the 
same word. Collapsing extra whitespace prevents formatting inconsistencies 
from interfering with tokenization.

I chose not to modernize spelling or standardize variations such as u/v or 
archaic spellings. While doing so might make the texts look more uniform, 
it could also erase meaningful stylistic or historical differences between 
authors. Since this assignment focuses on exploring patterns rather than rewriting 
the texts, I aimed to preserve the original language as much as possible.

```{r}
corpus_normalized <- corpus_raw %>%
  mutate(
    text = str_replace_all(text, "ſ", "s") %>%   # normalize long S
           str_replace_all("\\s+", " ") %>%     # collapse extra whitespace
           str_to_lower()                        # standardize case
  )

corpus_normalized
```

During tokenization, I also removed punctuation, standalone numbers, common 
stopwords (like “the” and “and”), and tokens containing obvious OCR errors 
(such as ● symbols). These elements do not carry meaningful content for this 
type of analysis and would otherwise introduce noise.

Overall, I used a minimal cleaning approach: correcting only what was necessary 
for accurate comparison while preserving the integrity of the original texts.


```{r}
tokens_df <- corpus_normalized %>%
  unnest_tokens(word, text)

```

```{r}

tokens_df <- tokens_df %>%
  anti_join(stop_words, by = "word") %>%         # remove stopwords
  filter(!str_detect(word, "^[0-9]+$")) %>%      # remove numbers
  filter(!str_detect(word, "●"))                 # remove OCR artifacts

# tokens_df
```

```{r}
#Check document length balance
tokens_df %>%
  count(doc_id, sort = TRUE)
```
```{r}
# Check most frequent words

tokens_df %>%
  count(word, sort = TRUE) %>%
  slice_max(n, n = 20)
```


```{r}
# Vocabulary size per document

tokens_df %>%
  group_by(doc_id) %>%
  summarise(unique_words = n_distinct(word))
```

```{r}
# Check for any non-alphabetic tokens

tokens_df %>%
  filter(str_detect(word, "[^a-z]")) %>%
  count(word, sort = TRUE)
```

These are real words and carry meaning, not OCR errors.

Approach 1: TF - IDF

```{r}
# Step 1: Document feature matrix

dfm_counts <- tokens_df %>%
  count(doc_id, word, sort = TRUE)

dfm_counts
```

```{r}

# Step 2: Compute TF–IDF weights

tf_idf_data <- dfm_counts %>%
  bind_tf_idf(word, doc_id, n) %>%
  arrange(desc(tf_idf))

tf_idf_data
```
```{r}

# Step 3: Top distinctive terms per document

top_terms <- tf_idf_data %>%
  group_by(doc_id) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup()

top_terms
```

Q1.Do some documents share distinctive vocabulary?

Yes. Even though TF–IDF measures distinctiveness relative to the entire corpus, several words appear as top TF–IDF terms in multiple documents. Words like monies, realm, commodities, exchange, and price show up as distinctive across more than one text.
This suggests that while each document has its own lexical profile, they still share a common conceptual vocabulary. In other words, TF–IDF highlights differences, but those differences are occurring within a shared thematic space. The documents are distinct, but not disconnected.

```{r}
top_terms <- tf_idf_data %>%
  group_by(doc_id) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup()
```

```{r}
colnames(top_terms)
```



```{r}
# Export one TF–IDF table per document

dir.create("tfidf_results", showWarnings = FALSE)

purrr::walk(names(doc_list), function(doc) {
  write_csv(doc_list[[doc]],
            file.path("tfidf_results",
                      paste0("tf_idf_", doc, ".csv")))
})
```


Q2.Are distinctive terms topical, rhetorical, or technical?
Most of the distinctive terms appear to be topical or technical, rather than purely rhetorical.

For example:
-> commodities, exchange, and price reflect economic or commercial terminology.
-> india and barbadoes indicate geographic or imperial contexts.
-> monies and realm reflect institutional or fiscal language.

This suggests that TF–IDF is capturing differences in subject matter and emphasis, rather than stylistic differences. The distinctiveness appears to reflect what each text is about, not just how it is written.

Q3.Is distinctiveness driven by noise or formatting?

There is little evidence that TF–IDF is being driven by noise. After removing stopwords, numbers, and OCR artifacts, the high TF–IDF terms are historically meaningful words rather than formatting errors or random fragments.
Some high TF–IDF words occur only once in a document (n = 1), which is expected because the inverse document frequency (IDF) component rewards rarity across the corpus. However, these words still appear to be substantively relevant (e.g., specific trade goods or place names), not technical glitches.
So overall, distinctiveness seems content-driven rather than noise-driven.

The TF–IDF results suggest that the corpus contains both shared vocabulary and document-specific emphasis. The documents appear to operate within a common economic and institutional framework, but each text highlights different aspects of that framework. This indicates structured variation rather than random divergence.
