---
title: "Data Exploration"
output: html_notebook
---

```{r}
library(tidyverse)
library(tidytext)
library(stringr)
library(widyr)

```


```{r}
getwd()
```
```{r}
list.files("texts", full.names = TRUE)
```
```{r}
read_file("texts/A06785.txt")
```


```{r}

files <- list.files("texts",
                    pattern = "\\.txt$",
                    full.names = TRUE)

corpus_raw <- tibble(
  doc_id = tools::file_path_sans_ext(basename(files)),
  text = map_chr(files, read_file)
)

print(corpus_raw)
```

```{r}
# Count words that contain "ſ" before replacement
sum(str_detect(corpus_raw$text, "ſ"))
```


Step 0: Normalization

In preparing the corpus for analysis, I adopted a minimal-intervention 
normalization strategy. The goal was to ensure lexical comparability across 
documents while preserving historically meaningful orthographic variation.


0.1) First, I replaced the Early Modern long S character (ſ) with the modern 
letter “s,” as required. Without this correction, words like “ſtate” and “state”
would be counted as different words, which would distort the analysis.

0.2) Second, I converted all text to lowercase and removed extra spaces. 
Lowercasing ensures that words such as “State” and “state” are treated as the 
same word. Collapsing extra whitespace prevents formatting inconsistencies 
from interfering with tokenization.

I chose not to modernize spelling or standardize variations such as u/v or 
archaic spellings. While doing so might make the texts look more uniform, 
it could also erase meaningful stylistic or historical differences between 
authors. Since this assignment focuses on exploring patterns rather than rewriting 
the texts, I aimed to preserve the original language as much as possible.

```{r}
corpus_normalized <- corpus_raw %>%
  mutate(
    text = str_replace_all(text, "ſ", "s") %>%   # normalize long S
           str_replace_all("\\s+", " ") %>%     # collapse extra whitespace
           str_to_lower()                        # standardize case
  )

corpus_normalized
```

During tokenization, I also removed punctuation, standalone numbers, common 
stopwords (like “the” and “and”), and tokens containing obvious OCR errors 
(such as ● symbols). These elements do not carry meaningful content for this 
type of analysis and would otherwise introduce noise.

Overall, I used a minimal cleaning approach: correcting only what was necessary 
for accurate comparison while preserving the integrity of the original texts.


```{r}
tokens_df <- corpus_normalized %>%
  unnest_tokens(word, text)

```

```{r}

tokens_df <- tokens_df %>%
  anti_join(stop_words, by = "word") %>%         # remove stopwords
  filter(!str_detect(word, "^[0-9]+$")) %>%      # remove numbers
  filter(!str_detect(word, "●"))                 # remove OCR artifacts

# tokens_df
```

```{r}
#Check document length balance
tokens_df %>%
  count(doc_id, sort = TRUE)
```
```{r}
# Check most frequent words

tokens_df %>%
  count(word, sort = TRUE) %>%
  slice_max(n, n = 20)
```


```{r}
# Vocabulary size per document

tokens_df %>%
  group_by(doc_id) %>%
  summarise(unique_words = n_distinct(word))
```

```{r}
# Check for any non-alphabetic tokens

tokens_df %>%
  filter(str_detect(word, "[^a-z]")) %>%
  count(word, sort = TRUE)
```

These are real words and carry meaning, not OCR errors.

APPROACH 1: TF - IDF

```{r}
# Step 1: Document feature matrix

dfm_counts <- tokens_df %>%
  count(doc_id, word, sort = TRUE)

dfm_counts
```

```{r}

# Step 2: Compute TF–IDF weights

tf_idf_data <- dfm_counts %>%
  bind_tf_idf(word, doc_id, n) %>%
  arrange(desc(tf_idf))

tf_idf_data
```
```{r}

# Step 3: Top distinctive terms per document

top_terms <- tf_idf_data %>%
  group_by(doc_id) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup()

top_terms
```


```{r}
top_terms <- tf_idf_data %>%
  group_by(doc_id) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup()
```

```{r}
colnames(top_terms)
```



```{r}
# Export one TF–IDF table per document

dir.create("tfidf_results", showWarnings = FALSE)

purrr::walk(names(doc_list), function(doc) {
  write_csv(doc_list[[doc]],
            file.path("tfidf_results",
                      paste0("tf_idf_", doc, ".csv")))
})
```

```{r}
# Create folder for TF–IDF distinctiveness results

dir.create("tfidf_distinctiveness", showWarnings = FALSE)

# Split results by document
doc_list <- split(top_terms, top_terms$doc_id)

# Export each document's distinctive terms
purrr::walk(names(doc_list), function(doc) {
  write_csv(doc_list[[doc]],
            file.path("tfidf_distinctiveness",
                      paste0("distinctive_terms_", doc, ".csv")))
})
```

```{r}
list.files("tfidf_distinctiveness")
```


APPROACH 2 - Pearson Correlation

```{r}
# Load required libraries

library(quanteda)
library(quanteda.textstats)
library(tidyverse)

```

```{r}
# Create quanteda corpus

corpus_quanteda <- corpus(corpus_normalized, text_field = "text")
#corpus_quanteda
```

```{r}
# Tokenize

tokens_quanteda <- tokens(
  corpus_quanteda,
  remove_punct = TRUE,
  remove_numbers = TRUE
)

```

```{r}
# Remove stop words
tokens_quanteda <- tokens_remove(tokens_quanteda, stopwords("en"))

#tokens_quanteda
```

```{r}
# Create DFM
dfm_counts <- dfm(tokens_quanteda)

dfm_counts
```
```{r}
dfm_trimmed <- dfm_trim(dfm_counts, min_termfreq = 5)

dfm_trimmed
```

```{r}
sim_r <- textstat_simil(dfm_trimmed,
                        margin = "documents",
                        method = "correlation")

r_mat <- as.matrix(sim_r)

r_mat <- round(r_mat, 3)

r_mat[1:5, 1:5]
```
```{r}
heat_df <- as.data.frame(r_mat) %>%
  rownames_to_column("doc_i") %>%
  pivot_longer(-doc_i,
               names_to = "doc_j",
               values_to = "r") %>%
  filter(doc_i != doc_j)

# Most similar pairs
most_similar <- heat_df %>%
  arrange(desc(r)) %>%
  distinct(pmin(doc_i, doc_j),
           pmax(doc_i, doc_j),
           .keep_all = TRUE) %>%
  slice_head(n = 2)

most_similar

```
```{r}
# Least similar pairs
least_similar <- heat_df %>%
  arrange(r) %>%
  distinct(pmin(doc_i, doc_j),
           pmax(doc_i, doc_j),
           .keep_all = TRUE) %>%
  slice_head(n = 2)



least_similar
```
```{r}
ggplot(heat_df,
       aes(x = doc_j,
           y = doc_i,
           fill = r)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_gradient2(
    low = "#2C7BB6",      # deep blue
    mid = "#FFFFBF",      # yellow
    high = "#D7191C",     # dark red
    midpoint = 0.5
  ) +
  labs(title = "Pearson Correlation Between Documents",
       x = NULL,
       y = NULL,
       fill = "Correlation") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```







APPROACH 3 -  Syntactic complexity profile

```{r}
library(udpipe)
library(dplyr)
library(stringr)
```

```{r}
model <- udpipe_download_model(language = "english")
ud_model <- udpipe_load_model(model$file_model)
```
```{r}
#Extract two texts 

text_subset <- corpus_normalized %>%
  filter(doc_id %in% c("A32827", "A93819"))
```

```{r}
parsed <- udpipe_annotate(ud_model,
                          x = text_subset$text,
                          doc_id = text_subset$doc_id)

parsed_df <- as.data.frame(parsed)
```

```{r}
# MLS

sentence_lengths <- parsed_df %>%
  group_by(doc_id, sentence_id) %>%
  summarise(words = n(), .groups = "drop") %>%
  group_by(doc_id) %>%
  summarise(MLS = mean(words))
```

```{r}
# Clauses per sentence

clauses <- parsed_df %>%
  filter(upos == "VERB") %>%
  group_by(doc_id) %>%
  summarise(total_clauses = n())
```

```{r}
sentences <- parsed_df %>%
  distinct(doc_id, sentence_id) %>%
  count(doc_id, name = "total_sentences")
```

```{r}
cs_ratio <- clauses %>%
  left_join(sentences, by = "doc_id") %>%
  mutate(C_per_S = total_clauses / total_sentences)
```

```{r}
colnames(parsed_df)
```



```{r}
# Dependent Clauses

dependent_clauses <- parsed_df %>%
  filter(dep_rel %in% c("ccomp", "xcomp", "advcl", "acl")) %>%
  group_by(doc_id) %>%
  summarise(dep_clauses = n())
```

```{r}
# Coordination

coordination <- parsed_df %>%
  filter(dep_rel == "conj") %>%
  group_by(doc_id) %>%
  summarise(coordination = n())
```

```{r}
# Complex Nominals

complex_nominals <- parsed_df %>%
  filter(dep_rel %in% c("amod", "nmod", "compound")) %>%
  group_by(doc_id) %>%
  summarise(complex_nominals = n())
```

```{r}
summary_table <- sentence_lengths %>%
  left_join(cs_ratio, by = "doc_id") %>%
  left_join(dependent_clauses, by = "doc_id") %>%
  left_join(coordination, by = "doc_id") %>%
  left_join(complex_nominals, by = "doc_id")

summary_table
```

```{r}
# Extract example sentences
example_sentences <- parsed_df %>%
  group_by(doc_id, sentence_id) %>%
  summarise(sentence = paste(token, collapse = " "),
            .groups = "drop") %>%
  group_by(doc_id) %>%
  slice_sample(n = 1)
```

```{r}
dir.create("syntactic_results", showWarnings = FALSE)

write_csv(example_sentences,
          file.path("syntactic_results",
                    "syntactic_example_sentences.csv"))
```

```{r}
list.files("syntactic_results")
```

```{r}
sentences_full <- parsed_df %>%
  group_by(doc_id, sentence_id) %>%
  summarise(
    sentence = paste(token, collapse = " "),
    word_count = n(),
    .groups = "drop"
  )

example_A32827 <- sentences_full %>%
  filter(doc_id == "A32827") %>%
  slice_max(word_count, n = 1)

example_A93819 <- sentences_full %>%
  filter(doc_id == "A93819") %>%
  slice_min(word_count, n = 1)

example_sentences <- bind_rows(example_A32827,
                               example_A93819)

example_sentences
```
```{r}
syntactic_measures <- summary_table %>%
  mutate(
    DC_per_S = dep_clauses / total_sentences,
    DC_per_C = dep_clauses / total_clauses,
    Coord_per_S = coordination / total_sentences,
    Coord_per_C = coordination / total_clauses,
    CN_per_S = complex_nominals / total_sentences,
    CN_per_C = complex_nominals / total_clauses
  ) %>%
  select(doc_id, MLS, C_per_S,
         DC_per_S, DC_per_C,
         Coord_per_S, Coord_per_C,
         CN_per_S, CN_per_C)
```

```{r}
syntactic_long <- syntactic_measures %>%
  pivot_longer(-doc_id,
               names_to = "measure",
               values_to = "value")
```

```{r}
ggplot(syntactic_long,
       aes(x = measure,
           y = value,
           fill = doc_id)) +
  geom_col(position = "dodge") +
  facet_wrap(~ measure, scales = "free") +
  scale_fill_manual(values = c("#4B3F72", "#1B998B")) +
  labs(title = "Syntactic Complexity Profile",
       x = NULL,
       y = "Value",
       fill = "Document") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        strip.text = element_text(face = "bold"))
```


